### 2021-02-27 newvariant.tex
Trying again to commit

The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---referred to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence.
Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility (which can be shared between both the wild type and the variant).
For simplicity, we do not explicitly model changes in the kernel.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
Assuming that the generation-interval distribution follows a gamma distribution, we can write the $r$--$\RR$ relationship in terms of the mean generation interval $\bar{G}_x$ and the effective dispersion $\kappa_x$, which captures the amount of variability in generation intervals:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
Here, the effective dispersion parameter $\kappa_x$ also corresponds to the squared coefficient of variation of the gamma distribution.
This gamma approximation framework has been used widely to model epidemic dynamics, including that of SARS-CoV-2.
We use this framework to understand how the inference of strength and speed of the variant depends on our assumptions about the underlying generation-interval distribution as well as the dynamics of the wild type.

@@

Typically, the speed of an epidemic $r_x$ can be measured directly from case data.
But when incidence of new variant is low (due to recent introduction) or unreliable (due to limited testing), estimating growth rate can be challenging.
During the current pandemic, many studies have focused on the characterizing the proportion of the new variant from genetic sequencing data---
when incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},\\
&= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
\end{align}
where the growth rate of the logistic equation $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.

>>>>

The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---referred to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence.
Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility. Many of these factors can affect between both the wild type and the variant in similar ways.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$); various gamma assumptions and approximations are widely used in epidemic modeling, including for modelsthat of SARS-CoV-2.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.

\jd{I like to talk about `effective dispersion' when we're getting into the details of distributions and how good the gamma assumption is. I think at this level we're OK with saying we approximate by a gamma and kappa is the squared CV.}

@@

Epidemic speed $r_x$ is often estimated directly from case data.
Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant.
This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wildtype viruses similarly.
When incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},
\\ &= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic. \jd{I now think it's easier to leave the second line in than to explain it in words.}
